% this is file `root', the root of the whole semantics
%\documentstyle[a4,12pt,twoside]{article}
%\documentclass[11pt,twoside]{article}
%\usepackage{a4wide}
\documentclass[12pt,twoside]{article}
\nonstopmode
\title{The Definition of Non-Standard ML\\
        (Syntax and Static Semantics) }
\date{Draft of \today}

\author{Claudio Russo \\
Laboratory for Foundations of Computer Science\\
Department of Computer Science\\
University of Edinburgh\\
\vspace{3ex}\\         
based on \\ 
{\sl The Definition of Standard ML} \\
{\sl Revised 1996} \\
by \\
{\sl   Robin Milner, Mads Tofte, Robert Harper and Dave MacQueen}\\
}
\include{mac} % macros
\makeindex
%

\includeonly{syncor, synmod, scope, statcor, statmod,app1}
% \includeonly{ statmod}

%for agfa: \voffset -12mm
\begin{document}
\pagestyle{empty}
\maketitle
\cleardoublepage
\pagestyle{plain}
\setcounter{page}{3}
\renewcommand{\thepage}{\roman{page}}


\include{preface}
\tableofcontents
\cleardoublepage
\pagestyle{headings}
\setcounter{page}{1}
\renewcommand{\thepage}{\arabic{page}}

\section{Introduction and Disclaimer}
This document, an extension of the definition of Standard ML, was never intended for publication. 
I drafted it during the design and implementation of Moscow ML and its extended Module system.
As far as I can recall, the latex sources were derived from a copy of the SML'90 Definition available at the LFCS,  manually updated to reflect the SML'97  revision
and then extended appropriately. The design only documents the changes to the syntax and static semantics of Standard ML and does not specify the required changes
to the dynamic semantics. Features added are higher-order functors (both applicative and generative), first-class modules and recursive modules  as well as minor
relaxations of the restrictions in Standard ML to make first-class modules more useful. The design aimed for backward compatibility with SML'97.
I am content with most of the design but, as it stands, the formalization of recursive modules is more complicated than it needs to be:
 the additional semantic objects called recursive structures are unnecessary and the system presented in my ICFP' 2001 paper is simpler.

{\it Claudio Russo, Microsoft Research, December 2014}. 


\include{intro}
\include{syncor}
\include{synmod}
\include{scope}
\include{statcor}
\include{statmod}
% \include{dyncor}
% \include{dynmod}
% \include{prog}
\appendix
\include{app1}

% \include{app2}
% \include{app3}
% \include{app4}
% \include{app5}
% \include{index}
\end{document}

 HOW TO REVISE THE INDEX 

The index is made using partly LaTex and partly an ML progam;
the latter is found on the file ``index.ml''. 

When LaTex is run on input ``root.tex'' it produces a file ``root.idx''.
Each line in this file is of the form

               \indexentry{idxkey}{pageref}

where idxkey is a key inserted precisely one place in the document 
(in the form of a LaTex command \index{idxkey})
and pageref is the page number of the page LaTex was printing
by the time it encountered the \index{idxkey} command.

A typical idxkey is 45.1 , which will occur in the TeX file 
somewhere near what produces page 45 in the final document. 
In fact, when the keys were first inserted in the TeX file, 
45.1 would be the first key on page 45, but as the document changes, 
one cannot get the pageref from the idxkey simple by taking the 
prefix of the idxkey up to the full stop.

There are many entries in the index that refer to the same
idxkey. Thus the number of idxkeys has been kept relatively
small, typically 2 or 3 pr page. The basic idea, then, is that
there is an ML program (on file ``index.ml'') which
associates entries in the final index with idxkeys by 
a sequence of expressions

              .......
              item ``handle'' [p``45.4'',``57.3''to``59.1'',p``78.2''];
              item ``{\\it happiness}'' [p''38.1''];
              ......
 
The program will first build a conversion table from idxkeys
(such as ``45.4'') to page references by reading thrugh
``root.idx''. Then it will evaluate all the item and subitem
expressions. The item function produces a line in thex
latex file (``index.tex'') using the conversion table.
Thus the above lines may produce

               ...........
               \item ``handle'' 46, 57--58, 78
               \item {\it happiness} 38
               ...........

If insertion of more text in the source files result in 
new page splits, then one should manually check that
the item expressions in ``index.ml'' refer to the
right idxkeys. It may be necessary to change some of the
lists in the item expressions and it may be desirable to
insert new \index commands in the source text. However,
if we for simplicity assume that we simply insert new
material corresponding to a new chapter (not affecting
the page splits in other chapters) then one would proceed
as follows:
First add new item expressions in the index.ml file corre-
sponding to new entries in the index (one will have new
\index commands in the new input, of course). Then one
runs latex on ``root.tex'' to produce the ``root.idx'' file. 
Then one runs index.ml (enter ML and type use ``index.ml'').
Then one runs latex on root again, this time giving the 
correct index.












